Deep ordinal classification based on cumulative link models

### 2019-05-23 paper/ieee.tex
Paper, tanh

	\section{Introduction}
	\label{sect:introduction}
	% The very first letter is a 2 line initial drop letter followed
	% by the rest of the first word in caps.
	% 
	% form to use if the first word consists of a single letter:
	% \IEEEPARstart{A}{demo} file is ....
	% 
	% form to use if you need the single drop letter followed by
	% normal text (unknown if ever used by the IEEE):
	% \IEEEPARstart{A}{}demo file is ....
	% 
	% Some journals put the first two words in caps:
	% \IEEEPARstart{T}{his demo} file is ....
	% 
	% Here we have the typical use of a "T" for an initial drop letter
	% and "HIS" in caps to complete the first word.
	\IEEEPARstart{D}{eep learning}, introduced by Yann Lecun~\cite{lecun2015deep}, combines multiple machine learning techniques and allows computational models that are composed of numerous processing layers to learn representations of data with various levels of abstraction. These methods have dramatically improved the state-of-the-art in many domains, such as image classification~\cite{cirecsan2012multi,he2016deep,krizhevsky2012imagenet}, speech recognition~\cite{hinton2012deep}, control problems~\cite{mnih2015human}, object detection~\cite{jiang2016speed,girshick2014rich}, privacy protection~\cite{yu2017iprivacy}, recovery of human pose~\cite{hong2015multimodal}, semantic segmentation~\cite{long2015fully} and image retrieval~\cite{li2015weakly}. Convolutional Neural Networks (CNN) are one of the types of deep networks that are designed to process data that comes in the form of multiple arrays. CNNs are appropriate for images, video, speech and audio processing, and they have been used extensively in the last years for automatic classification tasks~\cite{dong2014learning,sun2013deep,ronneberger2015u}. On image classification tasks, each colour channel is represented by a 2D array. In this case, convolutional layers extract the main features from the pixels of the images and, after that, a fully connected layer classify every sample based on its extracted features. At the output of the CNN, a softmax function provides the probabilities of the set of classes predefined in the model for classification tasks. These outputs are compared against the correct values.

	Ordinal classification problems are those classification tasks where labels are ordered, and there are different inter-classes importances for each pair of classes. This kind of problem can be treated as a nominal classification problem, but this discards the ordinal information. A better approach is to use specific methods that take the ordinality into account to improve the performance of the classification model. The Proportional Odds Model (POM)~\cite{mccullagh1980regression} is an ordinal alternative to the binary logistic regression. It belongs to a wider family of models called Cumulative Link Models (CLMs)~\cite{agresti2010analysis}. CLMs are inspired in the concept of a latent variable that is projected in a 1-dimensional space and a set of thresholds that divides that space into the different ordinal levels. This kind of models uses a link function which can be of different types, although the most common link function is the \texttt{logit}, which is used in POM. We explore different alternatives, as explained in depth in Section \ref{sect:ordinalproblem}.

	Ordinal models can also be applied to deep learning models. In the case of CNNs, the model projection used by the threshold model can be obtained from the last layer of the network. Given that we work with a 1-dimensional space, the last layer has only one neuron (projection of the pattern), and its value is used to classify the sample into the corresponding class according to the thresholds. Some previous works have used the \texttt{logit} in shallow neural networks~\cite{gutierrez2016ordinal}, but this strategy has not been considered for deep learning, and alternative link functions have not been evaluated. To further improve the results, we train these models by minimising an ordinal loss function based on the Weighted Kappa index \cite{de2018weighted}, instead of using the standard cross-entropy.

	In this paper, we propose the use of CLMs for deriving deep learning ordinal classifiers. An experimental study evaluating the three most common link functions is performed. Also, other parameters that can affect the training process and the model performance are studied, such as the learning rate of the optimization algorithm, the batch size, and their interaction. The nominal version of this model is used as a baseline for comparison. We contrast the results obtained with a statistical analysis to provide more robust conclusions. An ANOVA III test \cite{miller1997beyond} followed by a posthoc Tukey's test \cite{tukey1949comparing} is performed over $5$ runs of the experiments, because of the demands of computational time required to run a higher number of executions. The experiments are run using two different ordinal datasets: Diabetic Retinopathy~\cite{de2018weighted}, which contains high-resolution fundus images related with diabetes disease, and Adience~\cite{beckham2017unimodal}, which includes human faces images associated with an age range.

	This paper is organized as follows: in Section~\ref{sect:relatedwork}, we take a look at previous works related to this paper. Section~\ref{sect:ordinalproblem} present a formal description of an ordinal problem and CLMs. In Section~\ref{sect:experiments}, we describe the model, the experiments and the datasets used, while, in Section~\ref{sect:results}, we present the results obtained and the statistical analysis. Finally, Section~\ref{sect:conclusions} exposes the conclusions of this work.

	\section{Related works}
	\label{sect:relatedwork}
	There are many works related to the application and development of CNN models \cite{liu2017deep}, but few works focus on ordinal classification problems. The approaches proposed on those works are mainly related to 
	solving the ordinal problem as multiple binary sub-problems, using an ordinal loss function, solving the problem as a multi-class problem with constraints or simply using an ordinal evaluation metric. These works are described below:

	\vspace{0.2cm}
	$\bullet$ \textit{Solving the ordinal problem as multiple binary sub-problems.}

%		\textbf{(Binarización. Resolución de varios subproblemas.)}
	Z. Niu et al. \cite{niu2016ordinal} proposed a learning approach to address ordinal regression problems using convolutional neural networks. They divided the problem into a series of binary classification sub-problems and proposed a multiple output CNN optimization algorithm to collectively solve these classification sub-problems, taking into account the correlation between them.

@@

%		\textbf{(Binarización. Resolución de varios subproblemas.)}
	Y. Liu et al. \cite{liu2017deep} proposed a new approach which transforms the ordinal regression problem to binary classification sub-problems and use triplets with instances from different categories to train deep neural networks. In this way, high-level features describing the ordinal relationship are extracted automatically. Given that triplets must be generated, this approach is only recommended for small datasets.

@@

	\vspace{0.2cm}
	$\bullet$ \textit{Using an ordinal loss function.}

%		\textbf{(Función de coste ordinal)}
	J. de la Torre et al.~\cite{de2018weighted} proposed the use of a continuous version of the QWK metric as loss function for the optimization algorithm. They compared this cost function against the traditional log-loss function across three different databases, including the Diabetic Retinopathy database as the most complex one. They proved that their function could improve the results as it reduces overfitting and training time. Also, they checked the importance of hyper-parameter tuning.

%		\textbf{(Función de coste ordinal y distribución unimodal.)}
	C. Beckham and C. Pal~\cite{beckham2017unimodal} proposed a straightforward technique to constrain discrete ordinal probability distributions to be unimodal, via the use of the Poisson and binomial probability distributions. The parameters of these distributions were learnt by using a deep neural network. They evaluated this approach on two large ordinal image datasets, including the Adience dataset used in this paper, obtaining promising results. They also proposed a simple squared-error reformulation \cite{beckham2016simple} that was sensitive to class ordering.

@@

	\vspace{0.2cm}
	$\bullet$ \textit{Solving the problem as a multi-class problem with constraints.}

@@

	\vspace{0.2cm}
	$\bullet$ \textit{Simply using an ordinal evaluation metric.}

%		\textbf{(Métrica ordinal (MAE))}
	M. ALALI et al. \cite{alali2018multi} proposed a complex CNN architecture for solving Twitter Sentiment Classification as an ordinal problem. They checked that using average pooling preserves significant features that provide more expressiveness to ordinal scale. They didn't propose any method to include the ordinal information into the classifier but they tried to find the best CNN model architecture based on an ordinal metric.

	Adience dataset has been used in other works for human age estimation but most of them solved the problem as a multi-class problem instead of using the ordinal relation between classes. E. Eidinger \cite{eidinger2014age} presented an approach using support vector machines and neural networks. J.-C. Chen \cite{chen2016cascaded} proposed a coarse-to-fine strategy for deep CNNs, while G. Levi \cite{levi2015age} presented another convolutional network model for age estimation.

	However, none of the works mentioned have combined a CNN model with an unimodal probability distribution, without dividing the problem in multiple binary sub-problems and using a continuous loss function based on one of the most common metrics for ordinal classification (Weighted Kappa). This is the approach that will be proposed in this work for solving two separate ordinal problems.

@@

	\section{Cumulative Link Models (CLM)}
	\label{sect:ordinalproblem}
	An ordinal classification problem consists in predicting the label $y$ of an input vector $\mathbf{x}$, where $\mathbf{x} \in \mathcal{X} \subseteq \mathds{R}^K$ and $y \in \mathcal{Y}~=~\{\mathcal{C}_1, \mathcal{C}_2, ..., \mathcal{C}_Q\}$, i.e. $\mathbf{x}$ is in a K-dimensional input space, and $y$ is in a label space of $Q$ different labels. The objective of the ordinal problem is to find a function $r : \mathcal{X} \rightarrow \mathcal{Y}$ to predict the labels or categories of new patterns, given a training set of $N$ samples, $D = \{(\mathbf{x}_i, y_i), i = 1, ..., N\}$. Labels have a natural ordering in ordinal problems: $\mathcal{C}_1 \prec \mathcal{C}_2 \prec ... \prec \mathcal{C}_Q$. The order between labels gives us the possibility to compare two different elements of $\mathcal{Y}$ by using the relation $\prec$. This is not possible under the nominal classification setting. In regression (where $y \in \mathds{R}$), real values in $\mathds{R}$ can be ordered by the standard $<$ operator, but labels in ordinal regression ($y \in \mathcal{Y}$) do not carry metric information, i.e. the category serves as a qualitative indication of the pattern rather than a quantitative one.

	The Proportional Odds Model (POM) arises from a statistical background and is one of the first models designed explicitly for ordinal regression~\cite{mccullagh1980regression}. It dated back to 1980 and is a member of a wider family of models recognised as Cumulative Link Models (CLMs)~\cite{agresti2010analysis}. CLMs predict probabilities of groups of contiguous categories, taking the ordinal scale into account. In this way, cumulative probabilities $P(y \prec C_q |\mathbf{x})$ are estimated, which can be directly related to standard probabilities:
	\begin{align}
		\nonumber P(y \preceq \mathcal{C}_q | \mathbf{x}) &= P(y = \mathcal{C}_1 | \mathbf{x}) + ... + P(y = \mathcal{C}_q | \mathbf{x}),\\
		\nonumber P(y = \mathcal{C}_q | \mathbf{x}) &= P(y \preceq \mathcal{C}_q | \mathbf{x}) - P(y \preceq \mathcal{C}_{q-1} | \mathbf{x}),
	\end{align}
	with $q = 2, ..., Q-1$, and considering that $P(y = \mathcal{C}_1 | \mathbf{x}) = P(y \preceq \mathcal{C}_1 | \mathbf{x})$ and $P(y \preceq \mathcal{C}_Q | \mathbf{x}) = 1$.

	The model is inspired by the notion of a latent variable, where $f(\textbf{x})$ represents a one-dimensional mapping. The decision rule $r: \mathcal{X} \rightarrow \mathcal{Y}$ is not fitted directly, but stochastic ordering of space $\mathcal{X}$ is satisfied by the following general model form \cite{herbrich2000large}:
	\begin{equation}
		\nonumber
		g^{-1}(P(y \preceq \mathcal{C}_q | \mathbf{x})) = b_q - f(\mathbf{x}), \quad q = 1, ..., Q-1,
	\end{equation}
	where $g^{-1} : [0,1] \rightarrow (-\infty, +\infty)$ is a monotonic function often termed as the inverse link function, and $b_q$ is the threshold defined for class $\mathcal{C}_q$. Consider the latent variable $y^* = f(\mathbf{x})^* = f(\mathbf{x}) + \epsilon$, where $\epsilon$ is the random component of the error. The most common choice for the probability distribution of $\epsilon$ is the logistic function (which is the default function for POM). Label $\mathcal{C}_q$ is predicted if and only if $f(\mathbf{x}) \in [b_{q-1}, b_q]$, where the function $f$ and $\mathbf{b} = (b_0, b_1, ..., b_{Q-1}, b_Q)$ are to be determined from the data. It is assumed that $b_0 = -\infty$ and $b_Q = +\infty$, so the real line defined by $f(\textbf{x}), \textbf{x} \in \mathcal{X}$, is divided into $Q$ consecutive intervals. Each interval corresponds to a category. The constraints $b_1 \le b_2 \le ... \le b_{Q-1}$ ensures that $P(y \preceq \mathcal{C}_q | \mathbf{x})$ increases with $q$ \cite{mccullagh1980regression}.

@@

		\item The \textit{complementary log-log} (\texttt{clog-log}) takes a response that is restricted to the $(0,1)$ interval and converts it into something in the $(-\infty, +\infty)$ interval (like \texttt{logit} and \texttt{probit} transformations). The \texttt{clog-log} expression is:
		\begin{equation}
		\nonumber
		\begin{aligned}
		\log[-\log[1 - P(y \preceq \mathcal{C}_q | \mathbf{x})]] =b_q - f(\mathbf{x}),
		\end{aligned}
		\label{eq:cloglog}
		\end{equation}
		with $q = 1, ..., Q-1,$ that is:
		\begin{equation}
		\nonumber
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = 1 - e^{-e^{b_q - f(\mathbf{x})}}, \quad q = 1, ..., Q-1.
		\label{eq:cloglog2}
		\end{equation}
	\end{itemize}

@@

		\item For the \texttt{probit}:		
		\begin{equation}
		\nonumber
		\begin{aligned}
		p \equiv P(y \preceq \mathcal{C}_q | \mathbf{x}) &= \Phi(b_q - f(\mathbf{x})) =\\
		&= \int_{-\infty}^{b_q - f(\mathbf{x})} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2}x^2} \mathrm{d}x,
		\end{aligned}
		\end{equation}
		which leads to:
		\begin{align}
		\nonumber \text{probit}(p) & = \Phi^{-1}(p) = b_q - f(\mathbf{x}),\\
		\nonumber -\text{probit}(1-p) &= \Phi^{-1}(1-p) = -b_q + f(\mathbf{x}),
		\end{align}
		where:		
		\begin{equation}
		\nonumber
		1 - p = \int_{-\infty}^{-b_q + f(\mathbf{x})} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2}x^2} \mathrm{d}x,
		\end{equation}
		\begin{equation}
		\nonumber
		p = 1 - \int_{-\infty}^{-b_q + f(\mathbf{x})} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2}x^2} \mathrm{d}x.
		\end{equation}
	\end{enumerate}

	Unlike \texttt{logit} and \texttt{probit}, the \texttt{clog-log} link is asymmetrical. It is frequently used when the probability of an event is very small or very large. When the given data is not symmetric in the $[0,1]$ interval and increase slowly at small to moderate value but increases sharply near 1, the \texttt{logit} and \texttt{probit} models are inappropriate, while \texttt{clog-log} can lead to better results.

	This order is achieved by defining the first threshold and calculating the rest of thresholds from the first in the following form:	
	\begin{equation}
	\nonumber
	b_n = b_1 + \sum_{i=1}^{n-1} \alpha_i^2, \quad n = 2, ..., N,
	\end{equation}
	where $b_1$ and $\alpha_i$ are learnable parameters, and $N$ is the number of classes.

	\section{Weighted Kappa as loss function}
	\label{sect:wk}
	The Kappa index is a well-known metric that measures the agreement between two different raters. The Weighted Kappa (WK)~\cite{ben2008comparison} is based on the Kappa index and adds different weights to the different types of disagreements based on a weight matrix. It is useful to evaluate the performance in ordinal problems as it gives a higher weight to the errors that are further from the correct class. This metric is defined as follows:
	\begin{equation}
	\nonumber
	\text{QWK} = 1 - \frac{\sum\limits^N_{i,j} \omega_{i,j} O_{i,j}}{\sum\limits^N_{i,j} \omega_{i,j} E_{i,j}},
	\end{equation}
	where $N$ is the number of samples rated, $\omega$ is the penalization matrix (in this case, quadratic weights are considered), $O$ is the confusion matrix, $E_{ij} = \frac{O_{i\bullet} O_{\bullet j}}{N}$, $O_{i\bullet}$ is the sum of the $i\text{-th}$ row and $O_{\bullet j}$ is the sum of the $j\text{-th}$ column.

	The WK defined above cannot be used as a loss function for the optimization algorithm as it is not continuous. However, it can be redefined in terms of probabilities of the predictions:
	\begin{equation}
	\nonumber
	\text{QWK}_c = \frac{\sum\limits_{k=1}^N \sum\limits_{q=1}^Q \omega_{t_k, q} P(y = \mathcal{C}_q | \mathbf{x}_k)}{\sum\limits_{i=1}^Q \frac{N_i}{N} \sum\limits_{j=1}^Q ( \omega_{i,j} \sum\limits_{k=1}^N P(y = \mathcal{C}_j | \mathbf{x}_k))},
	\end{equation}
	where $\text{QWK}_c \in [0,2]$, $\mathbf{x}_k$ and $t_k$ are the input data and the real class of the $k$-th sample, $Q$ is the number of classes, $N$ is the number of samples, $N_i$ is the number of samples of the $i$-th class, $P(y = \mathcal{C}_q | \mathbf{x}_k)$ is the probability that the $k$-th sample belongs to class $\mathcal{C}_q$ (estimated using the CLM structure), and $\omega_{i,j}$ are the elements of the penalization matrix. In this case, $\omega_{i,j} = \frac{(i-j)^2}{(C-1)^2}$, where $\omega_{i,j} \in [0,1]$.

	This loss function can be minimized using a gradient descent based algorithm.

	\section{Experiments}
	\label{sect:experiments}
	\subsection{Data}
	In order to evaluate the different models, we make use of two ordinal datasets:

>>>>

	\section{Introduction}
	\label{sect:introduction}
	% The very first letter is a 2 line initial drop letter followed
	% by the rest of the first word in caps.
	% 
	% form to use if the first word consists of a single letter:
	% \IEEEPARstart{A}{demo} file is ....
	% 
	% form to use if you need the single drop letter followed by
	% normal text (unknown if ever used by the IEEE):
	% \IEEEPARstart{A}{}demo file is ....
	% 
	% Some journals put the first two words in caps:
	% \IEEEPARstart{T}{his demo} file is ....
	% 
	% Here we have the typical use of a "T" for an initial drop letter
	% and "HIS" in caps to complete the first word.
	\IEEEPARstart{D}{eep learning}, introduced by Yann Lecun~\cite{lecun2015deep}, combines multiple machine learning techniques and allows computational models that are composed of numerous processing layers to learn representations of data with various levels of abstraction. These methods have dramatically improved the state-of-the-art in many domains, such as image classification~\cite{cirecsan2012multi,he2016deep,krizhevsky2012imagenet}, speech recognition~\cite{hinton2012deep}, control problems~\cite{mnih2015human}, object detection~\cite{jiang2016speed,girshick2014rich}, privacy protection~\cite{yu2017iprivacy}, recovery of human pose~\cite{hong2015multimodal}, semantic segmentation~\cite{long2015fully} and image retrieval~\cite{li2015weakly}. Convolutional Neural Networks (CNN) are one of the types of deep networks that are designed to process data that come in the form of multiple arrays. CNNs are appropriate for images, video, speech and audio processing, and they have been used extensively in the last years for automatic classification tasks~\cite{dong2014learning,sun2013deep,ronneberger2015u}. For images, each colour channel is represented by a 2D array, and convolutional layers extract the main features from the pixels, and, after that, a fully connected layer classify every sample based on its extracted features. At the output of the CNN, a softmax function provides the probabilities of the set of classes predefined in the model for classification tasks. However, the softmax may not be the best option depending on the problem considered.

	Ordinal classification problems are those classification tasks where labels are ordered, and there are different inter-classes importances for each pair of categories. This kind of problem can be treated as nominal classification, but this discards the ordinal information. A better approach is to use specific methods that take the ordinality into account to improve the performance of the classification model. The Proportional Odds Model (POM)~\cite{mccullagh1980regression} is an ordinal alternative to the binary logistic regression. It belongs to a wider family of models called Cumulative Link Models (CLMs)~\cite{agresti2010analysis}. CLMs are inspired in the concept of a latent variable that is projected into a 1-dimensional space and a set of thresholds that divides the projection into the different ordinal levels. A link function needs to be specified, which can be of different types, although the most common option is the \texttt{logit}, which is used in POM. In this paper, we explore different existing alternatives, as explained in depth in Section \ref{sect:ordinalproblem}.

	In this paper, we propose the use of CLMs for deriving deep learning ordinal classifiers. In the case of CNNs, the model projection used by the threshold model can be obtained from the last layer of the network. Given that we work with a 1-dimensional space, the last layer would have only one neuron (projection of the pattern), and its value could be used to classify the sample into the corresponding class according to the thresholds. Some previous works have used the \texttt{logit} in shallow neural networks~\cite{gutierrez2016ordinal}, but this strategy has not been considered for deep learning, and alternative link functions have not been evaluated. To further improve the results, we train these models by minimising an ordinal loss function based on the Weighted Kappa index \cite{de2018weighted}, instead of using the standard cross-entropy.

	An experimental study evaluating the three most common link functions is performed. Also, other parameters that can affect the training process and the model performance are studied, such as the learning rate of the optimization algorithm, the batch size, and their interaction. The nominal version of this model is used as a baseline for comparison. We contrast the results obtained with a statistical analysis to provide more robust conclusions. An ANOVA III test \cite{miller1997beyond}, followed by a posthoc Tukey's test \cite{tukey1949comparing}, is performed over $5$ runs of the experiments, because of the demands of computational time required to run a higher number of executions. The experiments are run using two different ordinal datasets: Diabetic Retinopathy~\cite{de2018weighted}, which contains high-resolution fundus images related with diabetes disease, and Adience~\cite{beckham2017unimodal}, which includes human faces images associated with an age range.

	The paper is organized as follows: in Section~\ref{sect:relatedwork}, we analyse previous related works. Section~\ref{sect:modelProposal} presents a formal description of the proposal in this paper, which combines a CLM with an ordinal loss function. In Section~\ref{sect:experiments}, we describe the experiments and the datasets used, while, in Section~\ref{sect:results}, we present the results obtained and the statistical analysis. Finally, Section~\ref{sect:conclusions} exposes the conclusions of this work.

	\section{Related works}
	\label{sect:relatedwork}
	There are many works related to the application and development of CNN models \cite{liu2017deep}, but few works focus on ordinal classification problems. The existing deep ordinal approaches are mainly based on simply using an ordinal evaluation metric, on solving the ordinal problem as multiple binary sub-problems, on using an ordinal loss functions or on constraining the probability distribution of the output layer. These works are described in the following subsections.

	\subsection{Simply using an ordinal evaluation metric.}

	%\textbf{(Métrica ordinal (MAE))}
	M. ALALI et al. \cite{alali2018multi} proposed a complex CNN architecture for solving Twitter Sentiment Classification as an ordinal problem. They checked that using average pooling preserves significant features that provide more expressiveness to ordinal scale. They didn't propose any method to include the ordinal information into the classifier, but they tried to find the best CNN model architecture based on an ordinal metric.

	\subsection{Solving the ordinal problem as multiple binary sub-problems}

%		\textbf{(Binarización. Resolución de varios subproblemas.)}
	Z. Niu et al. \cite{niu2016ordinal} proposed a learning approach to address ordinal regression problems using CNNs. They divided the problem into a series of binary classification sub-problems and proposed a multiple output CNN optimization algorithm to collectively solve these classification sub-problems, taking into account the correlation between them.

@@

%		\textbf{(Binarización. Resolución de varios subproblemas.)}
	Y. Liu et al. \cite{liu2017deep} proposed a new approach which transforms the ordinal regression problem to binary classification sub-problems and use triplets with instances from different categories to train deep neural networks. In this way, high-level features describing the ordinal relationships are extracted automatically. Given that triplets must be generated, this approach is only recommended for small datasets.

@@

	In general, all these approaches increment the number of parameters to adjust, as several binary classifiers are simultaneously learnt.

	\subsection{Using an ordinal loss function}

%		\textbf{(Función de coste ordinal)}
	J. de la Torre et al.~\cite{de2018weighted} proposed the use of a continuous version of the quadratic weighted kappa (QWK) metric as loss function for the optimization algorithm. They compared this cost function against the traditional log-loss function using three different datasets, including the Diabetic Retinopathy database as the most complex one. They proved that their function could improve the results as it reduces overfitting and training time. Also, they checked the importance of hyper-parameter tuning.

@@

@@

	Although the use of these losses introduces the ordinality in model learning, the nature of the models remain nominal.

	\subsection{Unimodal probability distributions}

	%\textbf{(Función de coste ordinal y distribución unimodal.)}
	C. Beckham and C. Pal~\cite{beckham2017unimodal} proposed a straightforward technique to constrain discrete ordinal probability distributions to be unimodal, via the use of the Poisson and binomial probability distributions. The parameters of these distributions were learnt by using a deep neural network. They evaluated this approach on two large ordinal image datasets, including the Adience dataset used in this paper, obtaining promising results. They also included a simple squared-error reformulation \cite{beckham2016simple} that was sensitive to class ordering.

	This approach is the one most related to the CLMs considered in this paper. However, CLMs indirectly model a latent space together with the set of threshold separating the ordered classes, which provides a more flexible and interpretable approach to deep ordinal classification.

@@

	\section{Model proposal}
	\label{sect:modelProposal}

	Based on the previous analysis of the state-of-the-art, our proposal is to combine a flexible threshold model in the output layer (different forms of a CLM) with an ordinal loss function, in order to better introduce ordinal constraints during learning.

	\subsection{Cumulative Link Model (CLM)}
	\label{sect:ordinalproblem}
	An ordinal classification problem consists in predicting the label $y$ of an input vector $\mathbf{x}$, where $\mathbf{x} \in \mathcal{X} \subseteq \mathds{R}^K$ and $y \in \mathcal{Y}~=~\{\mathcal{C}_1, \mathcal{C}_2, ..., \mathcal{C}_Q\}$, i.e. $\mathbf{x}$ is in a $K$-dimensional input space, and $y$ is in a label space of $Q$ different labels. The objective in an ordinal problem is to find a function $r : \mathcal{X} \rightarrow \mathcal{Y}$ to predict the labels or categories of new patterns, given a training set of $N$ samples, $D = \{(\mathbf{x}_i, y_i), i = 1, ..., N\}$. Labels have a natural ordering in ordinal problems: $\mathcal{C}_1 \prec \mathcal{C}_2 \prec ... \prec \mathcal{C}_Q$. The order between labels gives us the possibility to compare two different elements of $\mathcal{Y}$ by using the relation $\prec$. This is not possible under the nominal classification setting. In regression (where $y \in \mathds{R}$), real values in $\mathds{R}$ can be ordered by the standard $<$ operator, but labels in ordinal regression ($y \in \mathcal{Y}$) do not carry metric information, i.e. the category serves as a qualitative indication of the pattern rather than a quantitative one.

	The Proportional Odds Model (POM) arises from a statistical background and is one of the first models designed explicitly for ordinal regression~\cite{mccullagh1980regression}. It dated back to 1980 and is a member of a wider family of models lately recognised as Cumulative Link Models (CLMs)~\cite{agresti2010analysis}. CLMs predict probabilities of groups of contiguous categories, taking the ordinal scale into account. In this way, cumulative probabilities $P(y \prec C_q |\mathbf{x})$ are estimated, which can be directly related to standard probabilities:
	\begin{align}
		\nonumber P(y \preceq \mathcal{C}_q | \mathbf{x}) &= P(y = \mathcal{C}_1 | \mathbf{x}) + ... + P(y = \mathcal{C}_q | \mathbf{x}),\\
		\nonumber P(y = \mathcal{C}_q | \mathbf{x}) &= P(y \preceq \mathcal{C}_q | \mathbf{x}) - P(y \preceq \mathcal{C}_{q-1} | \mathbf{x}),
	\end{align}
	with $q = 2, ..., Q-1$, and considering that $P(y = \mathcal{C}_1 | \mathbf{x}) = P(y \preceq \mathcal{C}_1 | \mathbf{x})$ and $P(y \preceq \mathcal{C}_Q | \mathbf{x}) = 1$.

	The model is inspired by the notion of a latent variable, where $f(\textbf{x})$ represents a one-dimensional mapping. The decision rule $r: \mathcal{X} \rightarrow \mathcal{Y}$ is not fitted directly, but stochastic ordering of space $\mathcal{X}$ is satisfied by the following general model form \cite{herbrich2000large}:
	\begin{equation}
		\nonumber
		g^{-1}(P(y \preceq \mathcal{C}_q | \mathbf{x})) = b_q - f(\mathbf{x}), \quad q = 1, ..., Q-1,
	\end{equation}
	where $g^{-1} : [0,1] \rightarrow (-\infty, +\infty)$ is a monotonic function often termed as the inverse link function, and $b_q$ is the threshold defined for class $\mathcal{C}_q$. Consider the latent variable $y^* = f(\mathbf{x})^* = f(\mathbf{x}) + \epsilon$, where $\epsilon$ is the random component of the error. The most common choice for the probability distribution of $\epsilon$ is the logistic function (which is the default function for POM). Label $\mathcal{C}_q$ is predicted if and only if $f(\mathbf{x}) \in [b_{q-1}, b_q]$, where the function $f$ and $\mathbf{b} = (b_0, b_1, ..., b_{Q-1}, b_Q)$ are to be determined from the data. It is assumed that $b_0 = -\infty$ and $b_Q = +\infty$, so the real line defined by $f(\textbf{x}), \textbf{x} \in \mathcal{X}$, is divided into $Q$ consecutive intervals. Each interval corresponds to a category. The constraints $b_1 \le b_2 \le ... \le b_{Q-1}$ ensure that $P(y \preceq \mathcal{C}_q | \mathbf{x})$ increases with $q$ \cite{mccullagh1980regression}.

@@

		\item The \texttt{clog-log} takes a response that is restricted to the $(0,1)$ interval and converts it into a value in the $(-\infty, +\infty)$ interval (like \texttt{logit} and \texttt{probit} transformations). The \texttt{clog-log} expression is:
		\begin{equation}
		\nonumber
		\begin{aligned}
		\log[-\log[1 - P(y \preceq \mathcal{C}_q | \mathbf{x})]] =b_q - f(\mathbf{x}),
		\end{aligned}
		\label{eq:cloglog}
		\end{equation}
		with $q = 1, ..., Q-1,$ that is:
		\begin{equation}
		\nonumber
		P(y \preceq \mathcal{C}_q | \mathbf{x}) = 1 - e^{-e^{b_q - f(\mathbf{x})}}, \quad q = 1, ..., Q-1.
		\label{eq:cloglog2}
		\end{equation}
	\end{itemize}

@@

		\item For the \texttt{probit}:		
		\begin{equation}
		\nonumber
		\begin{aligned}
		p \equiv P(y \preceq \mathcal{C}_q | \mathbf{x}) &= \Phi(b_q - f(\mathbf{x})) =\\
		&= \int_{-\infty}^{b_q - f(\mathbf{x})} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2}x^2} \mathrm{d}x,
		\end{aligned}
		\end{equation}
		which leads to:
		\begin{align}
		\nonumber \Phi^{-1}(p) & = \Phi^{-1}(p) = b_q - f(\mathbf{x}),\\
		\nonumber -\Phi^{-1}(1-p) &= \Phi^{-1}(1-p) = -b_q + f(\mathbf{x}),
		\end{align}
		where:		
		\begin{equation}
		\nonumber
		1 - p = \int_{-\infty}^{-b_q + f(\mathbf{x})} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2}x^2} \mathrm{d}x,
		\end{equation}
		\begin{equation}
		\nonumber
		p = 1 - \int_{-\infty}^{-b_q + f(\mathbf{x})} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2}x^2} \mathrm{d}x.
		\end{equation}
	\end{enumerate}

	Unlike \texttt{logit} and \texttt{probit}, the \texttt{clog-log} link is asymmetrical. In this way, when the given data is not symmetric in the $[0,1]$ interval and increase slowly at small to moderate value but increases sharply near 1, the \texttt{logit} and \texttt{probit} models are inappropriate, while \texttt{clog-log} can lead to better results.

	In this paper, the probabilistic structure of CLMs is proposed as a link function for deep convolutional neural networks. This can be achieved by defining a new type of output layer alternative to the standard softmax layer. In this way, the proposed output layer will transform the one-dimensional projection, previously denoted as $f(\mathbf{x})$, into a set of probabilities. $f(\mathbf{x})$ is estimated from a nonlinear transformation of the set of features learnt by the previous layers, $l(\mathbf{x})=f(\mathbf{x})$, where $\mathbf{x}$ is the pattern being evaluated and $l(\mathbf{x})$ is a latent representation of the pattern given by the output of a single neuron. In order to apply unconstrained optimizers while ensuring $b_1 \le b_2 \le ... \le b_{Q-1}$, we can redefine the thresholds. All of them can be derived from the first one in the following form:
	\begin{equation}
	b_q = b_1 + \sum_{q=1}^{q-1} \alpha_q^2, \quad q = 2, ..., Q,
	\end{equation}
	where $b_1$ and $\alpha_q$ are the learnable parameters, and $Q$ is the number of classes.

	\subsection{Continuous Quadratic Weighted Kappa (QWK) loss function}
	\label{sect:wk}

	In order to increase the performance of the deep ordinal model, the CLM structure in the output layer is combined with the continuous version of the QWK loss \cite{de2018weighted} function. The Kappa index is a well-known metric that measures the agreement between two different raters. The Weighted Kappa (WK)~\cite{ben2008comparison} is based on the Kappa index and adds different weights to the different types of disagreements based on a weight matrix. It is useful to evaluate the performance in ordinal problems, as it gives a higher weight to the errors that are further from the correct class. This metric is defined as follows:
	\begin{equation}
	\label{eq:qwk}
	\text{QWK} = 1 - \frac{\sum\limits^N_{i,j} \omega_{i,j} O_{i,j}}{\sum\limits^N_{i,j} \omega_{i,j} E_{i,j}},
	\end{equation}
	where $N$ is the number of samples rated, $\omega$ is the penalization matrix (in this case, quadratic weights are considered, $\omega_{i,j} = \frac{(i-j)^2}{(C-1)^2}$, $\omega_{i,j} \in [0,1]$), $O$ is the confusion matrix, $E_{ij} = \frac{O_{i\bullet} O_{\bullet j}}{N}$, $O_{i\bullet}$ is the sum of the $i\text{-th}$ row and $O_{\bullet j}$ is the sum of the $j\text{-th}$ column.

	The WK defined above cannot be used as a loss function for the optimization algorithm as it is not continuous. However, it has been previously redefined \cite{de2018weighted} in terms of probabilities of the predictions:
	\begin{equation}
	\nonumber
	\text{QWK}_c = \frac{\sum\limits_{k=1}^N \sum\limits_{q=1}^Q \omega_{t_k, q} P(y = \mathcal{C}_q | \mathbf{x}_k)}{\sum\limits_{i=1}^Q \frac{N_i}{N} \sum\limits_{j=1}^Q ( \omega_{i,j} \sum\limits_{k=1}^N P(y = \mathcal{C}_j | \mathbf{x}_k))},
	\end{equation}
	where $\text{QWK}_c \in [0,2]$, $\mathbf{x}_k$ and $t_k$ are the input data and the real class of the $k$-th sample, $Q$ is the number of classes, $N$ is the number of samples, $N_i$ is the number of samples of the $i$-th class, $P(y = \mathcal{C}_q | \mathbf{x}_k)$ is the probability that the $k$-th sample belongs to class $\mathcal{C}_q$ (estimated using the CLM structure), and $\omega_{i,j}$ are the elements of the penalization matrix ($\omega_{i,j} = \frac{(i-j)^2}{(C-1)^2}$). This loss function can be minimized using a gradient descent based algorithm.
